Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-03-16 _paper.rmd
Use RJH's jmlr template. Figure sizes and html consistency.

---
title: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
  - R
  - XAI
  - explanable artificial intelligence
  - black-box models
  - local explanations
  - grand tour
  - manual tour
  - radial tour
author:
- Nicholas Spyrison 
#^[Monash University, nicholas.spyrison@monash.edu, ORCiD: 0000-0002-8417-0212]
- Dianne Cook
- Kimbal Marriott
#^[ORCiD: 0000-0002-9813-0377]
  #- name: "Nicholas Spyrison"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 The increased predictive power of nonlinear models comes at the cost of interpretability of its terms.  This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction with _local explanations_, a point estimate of the linear feature importance in the vicinity of one instance. These can be considered linear projections and can be further explored to understand better the interactions between features used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) output. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr2e
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  pdf_document:
  ## bookdown::pdf_document2: allows \@ref(), but doesn't add authors....
  #bookdown::pdf_document2:
    #includes:
    #  in_header: template/jmlr2e.sty
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

```{r shapdistrbd, echo=F, fig.align='center', out.width="100%", fig.cap = "Illustration of SHAP values for a random forest model FIFA 2020 player wages from nine skill predictors. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) shows breakdown plots of three sequences of the features. The sequence of the variables impacts the magnitude of their attribution. Panel (b) shows the distribution of attribution for each feature across 25 sequences of predictors, with the mean displayed as a dot for each player. Reaction skills are important for both players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

@@

- a version of the cheem viewer shiny app can be directly accessed at [ebsmonash.shinyapps.io/cheem_initial/](https://ebsmonash.shinyapps.io/cheem_initial/). 
- the development version of the package is available at https://github.com/nspyrison/cheem, and 
- documentation of the package can be found at https://nspyrison.github.io/cheem/.

@@

(ref:casepenguins-cap) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of `fl` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at [vimeo.com/666431172](https://vimeo.com/666431172).)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins-cap)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

(ref:casechocolates-cap) Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the local explanation does not help understand the prediction for this instance. So, the contribution of Sugar is varied --- reducing it corresponds primarily with increasing Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this particular view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at [vimeo.com/666431143](https://vimeo.com/666431143).)
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates-cap)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

(ref:casechocolatesinverse-cap) Examining the local interpretation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange). In the attribution projection, the PI could be either milk or dark. Sodium and Fiber have the largest differences in attributed feature importance, with low values relative to other milk chocolates. The lack of importance attributed to these variables is suspected of contributing to the mistake, so the contribution of Sodium is varied. If Sodium had a larger contribution to the prediction (like in this view). the PI would look more like other milk chocolates. (A video of the tour animation can be found at [vimeo.com/666431148](https://vimeo.com/666431148).)
```{r casechocolatesinverse, echo=F, out.width="100%", fig.cap = "(ref:casechocolatesinverse-cap)"}
knitr::include_graphics("./figures/case_chocolates_inverse.png")
```

@@

(ref:casefifa-cap) Exploring the wages [euros] relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown at left, and it can be seen that this combination of features produces a view where Messi has very high predicted (and observed) wages. Defense (`def`) is the chosen feature to vary. It starts off very low, and as its contribution is increased (right plot) it can be seen that Messi's predicted wages decrease dramatically. The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that indeed, Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at [vimeo.com/666431163](https://vimeo.com/666431163).)
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa-cap)"}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

(ref:caseames-cap) Exploring an instance with a large residual as the PI from fitting sales price [USD] to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (`LtA`). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of Lot Area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at [vimeo.com/666431134](https://vimeo.com/666431134).)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames-cap)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader $\text{MI}^\text2$ lab group for the __DALEX__ ecosystem of __R__ and __Python__ packages. This research was supported by Australian Government Research Training Program (RTP) scholarships. Thanks to Jieyang Chong for helping proofread this article.

>>>>

---
title: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
shorttitle: "Cheem, Exploring Local Explanations"
author:
- familyname: Spyrison
  othernames: Nicholas
  address: Monash University\newline Australia\newline ORCiD 0000-0002-8417-0212
  email: Nicholas.Spyrison@monash.edu
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University\newline ORCiD 0000-0002-3813-7155
  email: DiCook@monash.edu
- familyname: Marriott
  othernames: Kimbal
  address: Monash University\newline ORCiD 0000-0002-9813-0377
  email: Kim.Marriott@monash.edu
# Nick ORCiD 0000-0002-8417-0212
# Di   ORCiD 0000-0002-3813-7155
# Kim  ORCiD 0000-0002-9813-0377
year: 2022
date_submitted: "MM/DD"
abstract: "The increased predictive power of nonlinear models comes at the cost of interpretability of its terms.  This trade-off has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models use predictors to arrive at a prediction with _local explanations_, a point estimate of the linear feature importance in the vicinity of one instance. These can be considered linear projections and can be further explored to understand better the interactions between features used to make predictions across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) output. The methods are implemented in the __R__ package __cheem__, available on CRAN."
keywords: "explainable artificial intelligence, nonlinear model interpretability, local explanations, grand tour, radial tour"
acknowledgment: "We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader MI 2 lab group for the DALEX ecosystem of R and Python packages. This research was supported by Australian Government Research Training Program (RTP) scholarships. Thanks to Jieyang Chong for helping proofread this article. The namesake, Cheem, refers to a fictional race of humanoid trees from Doctor Who lore. DALEX pulls on from that universe, and we initially apply tree SHAP explanations specific to tree-based models."
bibliography: cheem_paper.bib
classoption: 11pt,twoside
output:
  bookdown::pdf_document2:
    template: template_rjh/jmlr_template.tex
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    # includes:
    #   in_header: template_rjh/preamble.tex
    keep_tex: yes
    number_sections: yes
    #citation_package: natbib
    latex_engine: pdflatex
---

@@

```{r shapdistrbd, echo=F, fig.align='center', out.width="85%", fig.cap = "Illustration of SHAP values for a random forest model FIFA 2020 player wages from nine skill predictors. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) shows breakdown plots of three sequences of the features. The sequence of the variables impacts the magnitude of their attribution. Panel (b) shows the distribution of attribution for each feature across 25 sequences of predictors, with the mean displayed as a dot for each player. Reaction skills are important for both players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

@@

- a version of the cheem viewer shiny app can be directly accessed at https://ebsmonash.shinyapps.io/cheem_initial/. 
- the development version of the package is available at https://github.com/nspyrison/cheem, and 
- documentation of the package can be found at https://nspyrison.github.io/cheem/.

@@

(ref:casepenguins-cap) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of `fl` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at https://vimeo.com/666431172.)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins-cap)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

(ref:casechocolates-cap) Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the local explanation does not help understand the prediction for this instance. So, the contribution of Sugar is varied --- reducing it corresponds primarily with increasing Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this particular view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at https://vimeo.com/666431143.)
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates-cap)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

(ref:casechocolatesinverse-cap) Examining the local interpretation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange). In the attribution projection, the PI could be either milk or dark. Sodium and Fiber have the largest differences in attributed feature importance, with low values relative to other milk chocolates. The lack of importance attributed to these variables is suspected of contributing to the mistake, so the contribution of Sodium is varied. If Sodium had a larger contribution to the prediction (like in this view). the PI would look more like other milk chocolates. (A video of the tour animation can be found at https://vimeo.com/666431148.)
```{r casechocolatesinverse, echo=F, out.width="100%", fig.cap = "(ref:casechocolatesinverse-cap)"}
knitr::include_graphics("./figures/case_chocolates_inverse.png")
```

@@

(ref:casefifa-cap) Exploring the wages [euros] relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown at left, and it can be seen that this combination of features produces a view where Messi has very high predicted (and observed) wages. Defense (`def`) is the chosen feature to vary. It starts off very low, and as its contribution is increased (right plot) it can be seen that Messi's predicted wages decrease dramatically. The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that indeed, Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at https://vimeo.com/666431163.)
```{r casefifa, echo=F, out.width = "90%", fig.cap = "(ref:casefifa-cap)"}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

(ref:caseames-cap) Exploring an instance with a large residual as the PI from fitting sales price [USD] to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (`LtA`). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of Lot Area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at https://vimeo.com/666431134.)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames-cap)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader MI$^2$ lab group for the __DALEX__ ecosystem of __R__ and __Python__ packages. This research was supported by Australian Government Research Training Program (RTP) scholarships. Thanks to Jieyang Chong for helping proofread this article.
